---
title: "Deseq2LungCancer"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Start from loading the count matrices produced in other script
```{r message = FALSE, warning = FALSE}
Count_1 <- readRDS(file = "Count_1")
Count_2 <- readRDS(file = "Count_2")
Count_3 <- readRDS(file = "Count_3")
Count_4 <- readRDS(file = "Count_4")
Count_5 <- readRDS(file = "Count_5")
Count_6 <- readRDS(file = "Count_6")
Count_7 <- readRDS(file = "Count_7")
Count_8 <- readRDS(file = "Count_8")
Count_9 <- readRDS(file = "Count_9")
Count_10 <- readRDS(file = "Count_10")
Count_11 <- readRDS(file = "Count_11")
Count_12 <- readRDS(file = "Count_12")



```

```{r message = FALSE, warning = FALSE}
countdata <- data.frame(Count_1$counts, Count_2$counts, Count_3$counts, Count_4$counts, Count_5$counts, Count_6$counts, Count_7$counts, Count_8$counts, Count_9$counts, Count_10$counts, Count_11$counts, Count_12$counts)
```
```{r}
coldata <- read.csv(file = "Data/coldata.csv", row.names=1, stringsAsFactors=FALSE)
```

Here deseq2 object is produced
```{r message = FALSE, warning = FALSE}
library("dplyr")
library("ggplot2")
library("vsn")
library("DESeq2")
ddsMat <- DESeqDataSetFromMatrix(countData = countdata,
                                 colData = coldata,
                                 design = ~ Cell)
dds <- DESeq(ddsMat)

```


Compare different transformation methods for PCA
```{r message = FALSE, warning = FALSE}

vsd <- vst(ddsMat, blind = FALSE)
rld <- rlog(ddsMat, blind = FALSE)
ntd <- normTransform(ddsMat)

df <- bind_rows(
  as_data_frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"),
  as_data_frame(assay(rld)[, 1:2]) %>% mutate(transformation = "rlog"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst", "rlog")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  

```


Different PCA plots are produced

```{r message = FALSE, warning = FALSE,  fig.show="hold", out.width="50%"}
plotPCA(vsd, intgroup = "Cell", ntop = 500)
plotPCA(rld, intgroup = "Cell", ntop = 500)
plotPCA(ntd, intgroup = "Cell", ntop = 500)
```


 All of the PCA plots show a clear clustering of our samples. This confirms the samples have good quality


 SD against mean is calculated per type of transformation vsd results look better both with this plots and with PCA. vsd will be used to compare group distances in PCA environment
```{r message = FALSE, warning = FALSE,  fig.show="hold", out.width="50%"} 
meanSdPlot(assay(ntd))
meanSdPlot(assay(rld))
meanSdPlot(assay(vsd))

```


 Vsd appear to be the most accurate as we don't expect intra group variations

The following are heatmaps which give information ragarding the general distances between groups
```{r message = FALSE, warning = FALSE}
sampleDists <- dist(t(assay(vsd)))
library("pheatmap")
library("RColorBrewer")

sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <-  vsd@colData@listData[["Cell"]]
colnames(sampleDistMatrix) <- vsd@colData@listData[["Cell"]]
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```

Poisson distances
```{r message = FALSE, warning = FALSE}
library("PoiClaClu")
poisd <- PoissonDistance(t(counts(dds)))

samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- dds$Cell
colnames(samplePoisDistMatrix) <- dds$Cell
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors)
```



individual genes can be studied. Example provided 
```{r message = FALSE, warning = FALSE,  fig.show="hold", out.width="50%"}
#topGene <- rownames(res)[which.min(res$padj)]
#plotCounts(dds, gene = topGene, intgroup=c("Cell"))
plotCounts(dds, gene = "NFE2L2", intgroup=c("Cell"))
```

Heatmaps of Top Variable genes
```{r message = FALSE, warning = FALSE,  fig.show="hold", out.width="50%"}
library("genefilter")
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 20)

mat  <- assay(vsd)[ topVarGenes, ]
mat  <- mat - rowMeans(mat)
anno <- as.data.frame(colData(vsd)[, "Cell"])
pheatmap(mat)#, annotation_col = anno )
library( "gplots" )
library( "RColorBrewer" )
library( "genefilter" )
topVarGenes <- head( order( rowVars( assay(vsd) ), decreasing=TRUE ), 35 )

heatmap.2( assay(vsd)[ topVarGenes, ], scale="row", 
     trace="none", dendrogram="column", 
     col = colorRampPalette( rev(brewer.pal(9, "RdBu")) )(255))#,
     #ColSideColors = c( Control="gray", DPN="darkgreen", OHT="orange" )[
       # colData(vsd)$treatment ] )
```



Count box plot and dispersion plots. Box plot shows there is no outlier among our sampes. Dispersion plot is also typical. "the final estimates shrunk from the gene-wise estimates towards the fitted estimates. Some gene-wise estimates are flagged as outliers and not shrunk towards the fitted value, (this outlier detection is described in the manual page for estimateDispersionsMAP)"
```{r message = FALSE, warning = FALSE,  fig.show="hold", out.width="50%"}
par(mar=c(8,5,2,2))
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2)
plotDispEsts(dds)

```

(To be done on results
Rank of Cook wald statistic . We use Cook’s distance, which is a measure of how much the fitted coefficients would change if an individual sample were removed. This is carried ou as automatic detection of count outliers
will add later with FDR etc)

Calculation of DEG:plus log2fold change


```{r message = FALSE, warning = FALSE}
dds$Cell <- relevel(dds$Cell, ref = "A549")
dds <- DESeq(dds)
NCI.H460vsA549 <- results(dds, contrast = c("Cell", "NCI-H460", "A549"), alpha = 0.05)
NCI.H460vsA549shrunk <- lfcShrink(dds, coef =4, type = "apeglm")



Calu1vsA549 <-   results(dds, contrast = c("Cell", "Calu1", "A549"), alpha = 0.05)
Calu1vsA549shrunk <- lfcShrink(dds, coef =2, type = "apeglm")


NCIH358vsA549 <-   results(dds, contrast = c("Cell", "NCI-H358", "A549"), alpha = 0.05)
NCIH358vsA549shrunk <- lfcShrink(dds, coef =3, type = "apeglm")


dds$Cell <- relevel(dds$Cell, ref = "Calu1")
dds <- DESeq(dds)
NCIH460vsCalu1 <-   results(dds, contrast = c("Cell", "NCI-H460", "Calu1"), alpha = 0.05)
NCIH460vsCalu1shrunk <- lfcShrink(dds, coef =3, type = "apeglm")


NCIH358vsCalu1 <-   results(dds, contrast = c("Cell", "NCI-H358", "Calu1"), alpha = 0.05)
NCIH358vsCalu1shrunk <- lfcShrink(dds, coef =4, type = "apeglm")

dds$Cell <- relevel(dds$Cell, ref = "NCI-H358")
dds <- DESeq(dds)
NCIH358vsNCIH460 <-   results(dds, contrast = c("Cell", "NCI-H460", "NCI-H358"), alpha = 0.05)
NCIH358vsNCIH460shrunk <- lfcShrink(dds, coef =4, type = "apeglm")

```



will use shrinkage for now
Matrices
```{r message = FALSE, warning = FALSE}
DEG_NCI.H460vsA549shrunk <- as.data.frame(NCI.H460vsA549shrunk)  %>%  dplyr::filter(padj < 0.05 & log2FoldChange > 0.5)  %>%  dplyr::arrange(desc(log2FoldChange)) %>% write.csv(file = "DEG_NCI.H460vsA549shrunk.csv")

DEG_Calu1vsA549shrunk <- as.data.frame(Calu1vsA549shrunk)  %>%  dplyr::filter(padj < 0.05 & log2FoldChange > 0.5)  %>%  dplyr::arrange(desc(log2FoldChange)) %>% write.csv(file = "DEG_Calu1vsA549shrunk.csv")

DEG_NCIH358vsA549shrunk <- as.data.frame(NCIH358vsA549shrunk)  %>%  dplyr::filter(padj < 0.05 & log2FoldChange > 0.5)  %>%  dplyr::arrange(desc(log2FoldChange)) %>% write.csv(file = "DEG_NCIH358vsA549shrunk.csv")

DEG_NCIH460vsCalu1shrunk <- as.data.frame(NCIH460vsCalu1shrunk)  %>%  dplyr::filter(padj < 0.05 & log2FoldChange > 0.5)  %>%  dplyr::arrange(desc(log2FoldChange))%>% write.csv(file = "DEG_NCIH460vsCalu1shrunk.csv")

DEG_NCIH358vsCalu1shrunk <- as.data.frame(NCIH358vsCalu1shrunk)  %>%  dplyr::filter(padj < 0.05 & log2FoldChange > 0.5)  %>%  dplyr::arrange(desc(log2FoldChange)) %>% write.csv(file = "DEG_NCIH358vsCalu1shrunk.csv")

DEG_NCIH358vsNCIH460shrunk <- as.data.frame(NCIH358vsNCIH460shrunk)  %>%  dplyr::filter(padj < 0.05 & log2FoldChange > 0.5)  %>%  dplyr::arrange(desc(log2FoldChange)) %>% write.csv(file = "DEG_NCIH358vsNCIH460shrunk.csv")


```



```{r message = FALSE, warning = FALSE}
library(EnhancedVolcano)
vp1 <- EnhancedVolcano(NCI.H460vsA549,
lab = rownames(NCI.H460vsA549),
x = 'log2FoldChange',
y = 'padj', pCutoff = 0.05,
title = "NCI.H460vsA549")
print(vp1)
```

```{r message = FALSE, warning = FALSE}
library(EnhancedVolcano)
vp1 <- EnhancedVolcano(NCI.H460vsA549shrunk,
lab = rownames(NCI.H460vsA549shrunk),
x = 'log2FoldChange',
y = 'padj', pCutoff = 0.05,
title = "NCI.H460vsA549")
print(vp1)
```


Next will be Kegg 

```{r message = FALSE, warning = FALSE}
library(pathview)
library(gage)
library(gageData)
data(kegg.sets.hs)
data(sigmet.idx.hs)
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]
head(kegg.sets.hs, 3)
```
```{r  message = FALSE, warning = FALSE}
library(org.Hs.eg.db)
library(AnnotationDbi) 
NCI.H460vsA549shrunk$symbol = mapIds(org.Hs.eg.db,
                     keys=row.names(NCI.H460vsA549shrunk), 
                     column="SYMBOL",
                     keytype="SYMBOL",
                     multiVals="first")
NCI.H460vsA549shrunk$entrez = mapIds(org.Hs.eg.db,
                     keys=row.names(NCI.H460vsA549shrunk), 
                     column="ENTREZID",
                     keytype="SYMBOL",
                     multiVals="first")
NCI.H460vsA549shrunk$name =   mapIds(org.Hs.eg.db,
                     keys=row.names(NCI.H460vsA549shrunk), 
                     column="GENENAME",
                     keytype="SYMBOL",
                     multiVals="first")
NCI.H460vsA549shrunk$ENSEMBL =   mapIds(org.Hs.eg.db,
                     keys=row.names(NCI.H460vsA549shrunk), 
                     column="ENSEMBL",
                     keytype="SYMBOL",
                     multiVals="first")


head(NCI.H460vsA549shrunk, 10)
```

```{r message = FALSE, warning = FALSE}
foldchanges = NCI.H460vsA549shrunk$log2FoldChange
names(foldchanges) = NCI.H460vsA549shrunk$entrez
head(foldchanges)

```


```{r message = FALSE, warning = FALSE}
keggres = gage(foldchanges, gsets=kegg.sets.hs, same.dir=TRUE)
# Look at both up (greater), down (less), and statatistics.
lapply(keggres, head)

```



```{r message = FALSE, warning = FALSE}
keggrespathways = data.frame(id=rownames(keggres$greater), keggres$greater) %>% 
  tbl_df() %>% 
  filter(row_number()<=1000) %>% 
  .$id %>% 
  as.character()
#keggrespathways

keggresids = substr(keggrespathways, start=1, stop=8)
head(keggresids)

```

```{r message = FALSE, warning = FALSE}
# Define plotting function for applying later
plot_pathway = function(pid) pathview(gene.data=foldchanges, pathway.id=pid, species="hsa", new.signature=FALSE)
# plot multiple pathways (plots saved to disk and returns a throwaway list object)
tmp = sapply(keggresids, function(pid) pathview(gene.data=foldchanges, pathway.id=pid, species="hsa"))
```




```{r}
#tmp
```

```{r message = FALSE, warning = FALSE}
library(clusterProfiler)
library(dplyr)
universe <-  NCI.H460vsA549shrunk$symbol
NCI.H460vsA549shrunk <- as.data.frame(NCI.H460vsA549shrunk)
sigGenes <- dplyr::filter(NCI.H460vsA549shrunk, padj < 0.05, log2FoldChange >1) #& !is.na(A549_NCIH460$symbol)

enrich_go <- enrichGO(
  gene= sigGenes$symbol,
  OrgDb = org.Hs.eg.db,
  keyType = "SYMBOL",
  ont = "ALL",
  universe = universe,
  qvalueCutoff = 0.05,
  pvalueCutoff = 0.05,
  readable=TRUE
)

enrich_go_tidy <- enrich_go %>% 
  slot("result") %>% 
  tibble::as_tibble() 
enrich_go_tidy
pdf(file = "GOEgzample.pdf", width = 6, height = 8)
dotplot(enrich_go, showCategory=15)
dev.off()
#emapplot(enrich_go)
x2 <- enrichplot::pairwise_termsim(enrich_go)
emapplot(x2,cex_label_category = 0.6, group_legend = TRUE)
```

```{r}

```



